Sequential rank agreement methods for comparison of ranked lists

### 2018-02-28 latex/biosts3.tex
Shrunk to meet size requirements

  A typical situation where
  these questions arise is in high-dimensional genomics studies such
  as genome-wide association studies where several analysis methods
  (e.g., regression methods, lasso, random forest) can be used
  identify and rank millions of gene variants according to their
  association with the outcome. The importance (i.e., ranking) of each
  gene variant may vary from method to method and a consensus summary
  of agreement of the findings is needed to determine which gene
  variants to investigate more closely in subsequent validation
  studies. To minimiize expenses it is only of interest to consider
  gene variants that have high ranking across the different methods.
  Another situation where multiple ranked lists appear is in machine
  learning where the stability of the ranks produced by a
  ``black-box''-technique can be evaluted by bootstrapping the data
  repeatedly and comparing the ranks obtained from training the models
  on the bootstrapped data. Assessing which items are stable (i.e.,
  consistant across bootstrap samples) will help to weed out possible
  fluke findings.}

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists.  The general idea is to define agreement
based on the sequence of ranks from a subset of the first $d$ items in
each list. As agreement metric we adapt the limits of agreement known
from agreement between quantitative variables
\citep{alt:bland:1983,Carstensen2010} but any measure for agreement
could essentially be used. Our proposed approach allows us to compare
multiple lists simultaneously, it provides a dynamic measure of
agreement as a function of the depth in the lists, it places higher
weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement allows us to infer any changepoints, i.e., a
list depth where a substantial change in the agreement of the lists
occur but we also provide asymptotical and randomization-based
graphical tools to compare the observed rank agreement to the expected
agreement found in non-informative data.  In this sense our approach
is a combination and generalization of some of the ideas of
\added{\citet{iriz:etal:2005}}, \citet{Carterette2009}, and
\citet{Boulesteix2009}. \citet{Carterette2009} compares two rankings
based on the distance between them as measured by a multivariate
Gaussian distribution and the latter presents an overview of
approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. \added{\citet{iriz:etal:2005}
  proposes a plot which is a special case of our setting where the
  agreement metric is the overlap proportion}. We show the asymptotic
distribution of the endpoint of agreement and discuss how to infer the
distribution in small-sample situations using computational
methods. This enables us to make inferences about the endpoint of
agreement even for situations where there is no actual changepoint
present but just a gradual decline in agreement.

@@

Other recent approaches consider the intersection of lists as the
basis for a similarity measure \citep{iriz:etal:2005}. However, simple
intersection also places equal weights on all depths of the list and
therefore \citet{Fagin2003}, and \citet{Webber2010} proposed weighted
intersections which put more emphasis on the top of the lists.
% One proposal is denoted the average overlap (AO) and used for
% comparison later on. 
Specifically, \citet{Webber2010} define their rank-biased overlap
(RBO) by weighting with a converging series to ensure that the top is
weighted higher than the potentially non-informative bottom of the
lists.  It is possible to use the existing methods to calculate
agreement of lists until a given depth, i.e., limited to the $d$ items
of each list. However, the interpretation may not be straightforward,
especially in the case of more than two lists, and they may not
accommodate partial rankings.

@@

The manuscript is organized as follows: The next section defines
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we discuss approaches to
evaluate the results obtained from sequential rank agreement. Finally
we apply the sequential rank agreement to two Danish ovarian cancer
studies and compare our method to the method of \citet{hall:schi:2012}
in a small sample simulation study before we discuss the findings
along with possible extensions. The approaches presented in this
manuscript are available in the R package \texttt{SuperRanker} which
can be found on CRAN.
% chunk 1

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$.  Panels (a) and (b) of Table~\ref{tab:example}
show a schematic example of these mappings. Thus if
$R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked first in list $l$
and similarly $R_l(X_{34})=1$.

In all what follows we consider a fixed set of items and consider the
ranking function to be a random variable. Thus, let
$R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $1/P!$ to each of the
$P!$ different possible rankings.

The agreement among the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used in method comparison studies to compute the limits of
agreement \citep{alt:bland:1983}.

@@

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{equation}
\added{  \textrm{sra}(d) = \left\{\begin{array}{cl}\frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)  & \text{when $|S(d)|>0$} \\ 0 & \text{otherwise}\end{array}  \right.}\label{def:sra}
\end{equation}
where $|S(d)|$ is the cardinality of the set $S(d)$. \added{As stated, we are only interested in sra($d$) when $|S(d)|>0$ so the value 0 above is somewhat arbitrary.}
The empirical counterpart when $|S(d)|>0$ is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}.\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero when $|S(d)|>0$ suggest that the
lists agree on the rankings while larger values suggest
disagreement. \added{If $|S(d)|=0$ then no items were sufficiently
  frequent among the observed lists and we can conclude that the lists
  do not agree above the threshold $\varepsilon$.  } If the ranked
lists are identical then the value of sequential rank agreement will
be zero for all values of $d$. %position or depths $d$ down the lists.

@@

As an example consider the data by \citet{Golub1999} (also found in
\citet{Dudoit2002}) where 3051 gene expression values measured on 38
tumor mRNA samples were used to classify between acute lymphoblastic
leukemia (ALL) and acute myeloid leukemia (AML). Several possible
analysis methods can be used for this for example marginal unequal
variances two-sample $t$ tests, marginal logistic regression analyses,
elastic net logistic regression \citep{friedman2010regularization},
and marginal maximum information content correlations (MIC)
\citep{Reshef2011}, and we would like to identify a set of genes that
consistently are most likely to be associated to leukemia.

@@

The sequential rank agreement curve (using $\varepsilon=0$ to decide
that an item should be present in just a single list before it is
included) shown in the left plot of Figure~\ref{fig:example1} show the
average distance in ranks for the genes considered among the first $d$
positions. Not surprisingly, the sequential rank agreement is better
towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. It is clear from the curve in Figure~\ref{fig:example1} that
there is a substantial deterioration in agreement (higher sra) after
depth 5. Thus, if we were to restrict attention to a small set of
predictors then our prime focus would be on the items found among the
top-5 lists from Table~\ref{tab1}). The choice of which depth until
which we think the lists agree can be chosen either from a
pre-specified threshold for an acceptable difference in rankings or
from a pre-specified item set size. A changepoint analysis on the
sequential rank agreement would be able to identify depths where a
substantial increase/change in rank agreement occurs and would be
another way to identify sets of items that share agreement among the
lists if a pre-specified acceptable rank agreement threshold is not
given.}

@@

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
Generally, changes in the level of rank agreement suggest that there
are sets of items that are ranked similarly in all lists while other
items constitute set(s) that have been assigned vastly different
ranks. When the lists are likely to agree on a few top ranked items
then the sequential rank agreement curve will start at a low level and
then increase until it levels off exactly as is seen in
Figure~\ref{fig:example1}.

@@

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. One realization of the $L$ rankings
of the set $X$ is obtained by randomizing the missing items of each
list. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~\ref{sra-algorithm}.

@@

Alternatively, we can remove the restriction of the independence among
the methods used to generate the $L$ ranked lists and only require
that there is no information contained in the rankings but that the
rankings are all based on applying the method/approaches to the same
data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

The incomplete data also suggests that there may be at most 1 or 2
ranked items towards the top of the lists that yield a result better
than what would be expected (the bottom-right plot). Not surprisingly,
the sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify (and rank) similar predictors even if there
are only spurious associations.

\added{It is important to stress that neither $H_0$ nor
  $\widetilde H_0$ are related to questions regarding the association
  between the outcome and the predictors in the dataset. Both
  hypotheses are purely considering how the rankings agree in a
  situation where there is no relevant information available in the
  data used for creating the rankings. It is also worth pointing out,
  that if the lists are short ($P$ low) and there are few lists ($L$
  low) then the number of possible different permutations under the
  null is small and the $p$ value obtained may be fluctuating if the
  number of permutations is small. We have found that a number of
  permutations over 500 works well for smaller samples. The number of
  permutations can be lowered for larger data sizes.}

@@

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from a
  probability distribution $Q$ on the set of lists $\Pi$. Then,
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\begin{proof}
See Appendix A in the supplementary material.%\ref{sec:appA}.
\end{proof}
\end{theorem}

We now define the changepoint as the first crossing point of the
sequential rank agreement and a threshold function
$q\colon\,\{1,\ldots,P\} \mapsto \mathbb{R}_{\geq 0}$. The values of
$q$ could be a deterministic constant or, % in the experimental setting,
for example, the limits-of-agreement obtained in randomly permuted
lists corresponding to the null-hypothesis in equation
(\ref{eq:permutationHypothesis}). We define the superlevel set of the
sequential rank agreement with respect to $q$ as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
A changepoint $d^\ast(q)$ in the list agreement is then defined by the position
\begin{equation}
  d^\ast(q) =
  \begin{cases}
    \inf(\mathcal{L}(q)) & |\mathcal{L}(q)| > 0\\
  P & |\mathcal{L}(q)| = 0
  \end{cases}
\end{equation}
corresponding to the first list depth where the sequential rank
agreement exceeds the threshold if such a position exists. Otherwise,
the full list is in agreement according to $q$ and the changepoint is
set to the full length of the lists. The empirical superlevel set is
similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where we allow the threshold function to depend on the sample size as
well. The estimated changepoint is therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated changepoint, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a positive threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then, $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\begin{proof}
See Appendix B in supplementary material.%\ref{sec:appB}.
\end{proof}
\end{corollary}

@@

The MALOVA study is a multidisciplinary Danish study on ovarian cancer
\citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999. For the purpose of illustration we use a random
sub-sample of $119$ patients with a total of $58$ patients with
malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another
multidisciplinary Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\citep{bertelsen1991protocol}. The study aimed to continuously enroll
all patients that were referred to surgery of an ovarian tumor
clinically suspected to be cancer during the period from 1984 to
1990. Similarly, we use a random sub-sample from the DACOVA study of
$113$ patients with a total of $54$ malignant ovarian cancers and $59$
benign ovarian tumors/gynecologic disorders.

@@

In both scenarios we use the MALOVA data to train the statistical
models, and in both situations the agreements are assessed with
respect to perturbations of the training data in the following
manner. We repeatedly draw a random sub-sample (without replication)
consisting of 90\% of the MALOVA observations and train the four
models on each sub-sample. We use 1000 iterations for the sub-sampling
procedure.

The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to our data to investigate their effect on the agreements.
Note also that the Lasso models produce incomplete lists depending on
the value of the penalty parameter.

@@

The right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
600. At deeper list depths all agreement curves are approximately constant.
%
As expected, most of the sequential rank agreement curves start low,
indicating good agreement, followed by an increase until they
approximately become constant. This has the interpretation that the
agreement across the different sub-samples is higher in the top as
compared to the tail of the lists for all these classification
methods. The changepoints where the curves become approximately
constant are the list depths where the ranks of the remaining items
become close to uniformly random.

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we
see that a large proportion of them are numerically very close to zero
and have almost equal absolute value. This is a general feature of the
Ridge models in this dataset and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient
values below a given threshold thereby introducing an artificial
incompletion of the lists.  For the Ridge models tuned with the
deviance criterion, Figure \ref{fig:app2} (left) shows the the
sequential rank agreement where for each of the 1000 trained models
the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated using Algorithm
\ref{sra-algorithm} with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (right panel) is shown for
comparison.  Even though the number of predictors with missing ranks is very
small compared to the total number of predictors, the effect on the
sequential rank agreement is substantial and with the artificial
censoring the shape of the curves is as expected, starting low and
then increasing.

Looking at the agreement curves for the Lasso models in Figure
\ref{fig:app1} (right) we clearly see the effect of the sparsity
inducing penalization giving rise to incomplete lists. These curves
were similarly calculated using \ref{sra-algorithm} and $1000$ random
permutations. Under the deviance optimization criterion the median
number of non-zero coefficients was 33 (range 16 to 50) and for the
class accuracy criterion 14 (range 4 to 56). These values correspond
to the list depths where the agreement curves become constant as a
result of the subsequent censoring.

>>>>

  A typical situation where these questions arise is in
  high-dimensional genomics studies such as genome-wide association
  studies where several analysis methods (e.g., regression methods,
  lasso, random forest) can be used identify and rank millions of gene
  variants according to their association with the outcome. The
  ranking of each gene variant may vary from method to method and a
  consensus summary of agreement of the findings is needed to
  determine which gene variants to investigate more closely in
  subsequent validation studies. To minimiize expenses it is only of
  interest to consider gene variants that have high ranking across the
  different methods.  Multiple ranked lists also appear in machine
  learning where the stability of the ranks produced by a
  ``black-box''-technique can be evaluted by bootstrapping the
  data. Assessing which items are stable (i.e., consistant rankings
  across bootstrap samples) will help to weed out possible fluke
  findings.}

In this article we introduce sequential rank agreement for measuring
agreement among ranked lists.  The general idea is to define agreement
based on the sequence of ranks from a subset of the first $d$ items in
each list. As agreement metric we adapt the limits of agreement known
from agreement between quantitative variables
\citep{alt:bland:1983} but any measure for agreement
could essentially be used. Our proposed approach allows us to compare
multiple lists simultaneously, it provides a dynamic measure of
agreement as a function of the depth in the lists, it places higher
weight on items at the top of the list, it accommodates partially
observed lists of varying lengths, and has a natural interpretation
that directly relates to the ranks.  Graphical illustration of
sequential rank agreement allows us to infer any changepoints, i.e., a
list depth where a substantial change in the agreement of the lists
occur but we also provide asymptotical and randomization-based
graphical tools to compare the observed rank agreement to the expected
agreement found in non-informative data.  In this sense our approach
is a combination and generalization of some of the ideas of
\added{\citet{iriz:etal:2005}}, \citet{Carterette2009}, and
\citet{Boulesteix2009}. \citet{Carterette2009} compares two rankings
based on the distance between them as measured by a multivariate
Gaussian distribution and the latter presents an overview of
approaches for aggregation of ranked lists including bootstrap and
leave-one-out jackknife approaches. \added{\citet{iriz:etal:2005}
  proposes a plot based on the intersection of lists which is a
  special case of our setting where the agreement metric is the
  overlap proportion}. However, simple intersection also places equal
weights on all depths of the list and therefore \citet{Fagin2003}, and
\citet{Webber2010} proposed weighted intersections which put more
emphasis on the top of the lists.  Specifically, \citet{Webber2010}
define their rank-biased overlap (RBO) by weighting with a converging
series to ensure that the top is weighted higher than the potentially
non-informative bottom of the lists.  It is possible to use the
existing methods to calculate agreement of lists until a given depth,
i.e., limited to the $d$ items of each list. However, the
interpretation may not be straightforward, especially in the case of
more than two lists, and they may not accommodate partial rankings.

@@

@@

The manuscript is organized as follows: The next section defines
sequential rank agreement for multiple ranked lists and discuss how to
handle incomplete lists. In section 3 we show the asymptotic
distribution of the endpoint of agreement and discuss approaches to
evaluate the results obtained from sequential rank agreement.
%infer the
%distribution in small-sample situations using computational
%methods. This enables us to make inferences about the endpoint of
%agreement even for situations where there is no actual changepoint
%present but just a gradual decline in agreement.
Finally we apply the sequential rank agreement to two Danish ovarian
cancer studies and compare our method to the method of
\citet{hall:schi:2012} in a small sample simulation study before we
discuss the findings and possible extensions. The approaches
presented are implemented in the R package \texttt{SuperRanker}  found on CRAN.
% chunk 1

@@

Consider a set of $P$ different items $X=\{X_1,\dots,X_P\}$ and a
ranking function $R: \{X_1,\dots,X_P\}\to \{1,\dots,P\}$, such that
$R(X_p)$ is the rank of item $X_p$. The inverse mapping $R^{-1}$ gives
the item $R^{-1}(r)$ that was assigned to rank $r\in\{1,\dots,P\}$. An
ordered list is the realization of a ranking function $R$ applied to
the set of items $X$.  Panels (a) and (b) of Table~\ref{tab:example}
show a schematic example of these mappings. Thus if
$R_l^{-1}(1)=X_{34}$ then item $X_{34}$ is ranked first in list $l$
and similarly $R_l(X_{34})=1$.
% 
In all what follows we consider a fixed set of items and consider the
ranking function to be a random variable. Thus, let
$R_1(X),\dots,R_L(X)$, $L\geq2$, be a sample of $L$ independent
identically distributed draws from an unknown probability distribution
function $Q$. One aim is then to test how much $Q$ resembles the
uniform distribution which assigns probability $1/P!$ to each of the
$P!$ different possible rankings.

The agreement among the lists regarding the rank given to an item $X_p$
can be measured by the variance across the lists
\begin{align}
  A(X_p) &= \E_Q\left[\left(R(X_p)-\E_Q R(X_p)\right)^2\right]\\
   &= \sum_{r\in\Pi} \left(r(X_p)-\E_Q R(X_p)\right)^2 Q(r),\nn
\end{align}
where $\Pi$ is the set of all permutations of $X$, $Q$ is a
probability mass function on $\Pi$, and
$\E_Q R(X_p)=\sum_{r\in\Pi} r(X_p)Q(r)$. The empirical counterpart is
\begin{align}
\widehat{A}_L(X_p) = \frac{1}{L-1}\sum_{i=1}^L (R_i(X_p) - \overline{R}_L(X_p))^2, \quad \overline{R}_L(X_p) = \frac{1}{L}\sum_{i=1}^L R_i(X_p).\label{eq:empDistance}
\end{align}
For each item, the function $\widehat{A}_L$ has an interpretation as the expected
Euclidean distance of the individual rankings from the expected
ranking over the $L$ lists, and it corresponds to the same measure
that is used to compute the limits of
agreement \citep{alt:bland:1983}.

@@

We define the \emph{sequential rank agreement} as the weighted
expected agreement of the items found in the set $S(d)$:
\begin{equation}
\added{  \textrm{sra}(d) = \left\{\begin{array}{cl}\frac{1}{|S(d)|}\sum_{p \in S(d)}A(X_p)  & \text{when $|S(d)|>0$} \\ 0 & \text{otherwise}\end{array}  \right.}\label{def:sra}
\end{equation}
where $|S(d)|$ is the cardinality of the set $S(d)$. \added{As stated, we are only interested in sra($d$) when $|S(d)|>0$.}
The empirical counterpart when $|S(d)|>0$ is equivalently given by
\begin{equation}
  \widehat{\textrm{sra}}_L(d) = \frac{\sum_{\{p \in \widehat{S}_{L}(d)\}}(L-1)\widehat{A}_L(X_p)}{(L-1)|\widehat {S}_{L}(d)|}.\label{def:sraEst}
\end{equation}

Values of \textrm{sra} close to zero when $|S(d)|>0$ suggest that the
lists agree on the rankings while larger values suggest
disagreement. \added{If $|S(d)|=0$ then no items were sufficiently
  frequent among the observed lists and we can conclude that the lists
  do not agree above the threshold $\varepsilon$.  } The sequential rank agreement will be zero for all values of $d$
when the ranked
lists are identical.
 %position or depths $d$ down the lists.

@@

As an example consider the data by \citet{Golub1999} (found in
\citet{Dudoit2002}) where 3051 gene expression values measured on 38
tumor mRNA samples were used to classify between acute lymphoblastic
leukemia (ALL) and acute myeloid leukemia (AML). Several 
analysis methods are possible for these data for example marginal unequal
variances two-sample $t$ tests, marginal logistic regression analyses,
elastic net logistic regression \citep{friedman2010regularization},
and marginal maximum information content correlations (MIC)
\citep{Reshef2011}, and we want to identify a set of genes that
consistently are most likely to be associated to leukemia.

@@

The sequential rank agreement curve (using $\varepsilon=0$ to decide
that an item should be present in just a single list before it is
included) shown in the left plot of Figure~\ref{fig:example1} show the
average distance in ranks for the genes considered among the first $d$
positions. Not surprisingly, the sequential rank agreement is better
towards the top of the lists (smaller values on the $y$ axis
corresponds to \emph{better} agreement) than towards the bottom of the
lists. Figure~\ref{fig:example1} shows a substantial deterioration in agreement (higher sra) after
depth 5. Thus, if we were to restrict attention to a small set of
predictors then our prime focus would be on the items found among the
top-5 lists from Table~\ref{tab1}). The choice of which depth until
which we think the lists agree can be chosen either from a
pre-specified threshold for an acceptable difference in rankings or
from a pre-specified item set size. A changepoint analysis on the
sequential rank agreement would be able to identify depths where a
substantial increase/change in rank agreement occurs and would be
another way to identify sets of items that share agreement among the
lists if a pre-specified acceptable rank agreement threshold is not
given.}

@@

% If the lists agree on all items then the sequential rank agreement is
% zero for all depths. 
Generally, changes in the level of rank agreement suggest that there
are sets of items that are ranked similarly in all lists while other
items constitute set(s) with vastly different
ranks. When the lists are likely to agree on a few top ranked items
then the sequential rank agreement curve will start low and
then increase until it levels off exactly as is seen in
Figure~\ref{fig:example1}.

@@

Instead we randomize the ranks $\{d_{l}+1,\dots,P\}$ to the items that
do not occur in list $\Lambda_l$. One realization of the $L$ rankings
of the set $X$ is obtained by randomizing the missing items of each
list. By randomizing a large number of times we can compute
\eqref{def:sra} for each realization, and then compute the sequential
rank agreement as the pointwise (for each depth) average of the rank
agreements. The algorithm is described in detail in
Algorithm~C in the appendix.

@@

Alternatively, we can remove the restriction of the independence among
the methods used to generate the $L$ ranked lists under the null
hypothesis that any association to the outcome is removed in the
data.

%and only require
%that there is no information contained in the rankings but that the
%rankings are all based on applying the method/approaches to the same
%data
\begin{eqnarray*}
\widetilde H_0 & :&  \text{The list rankings are based on data that contain}\\
& &   \text{no association to the outcome.}
\end{eqnarray*}

@@

The incomplete data also suggests that there may be at most 1 or 2
ranked items towards the top of the lists that yield a result better
than what would be expected (the bottom-right plot). Not surprisingly,
the sequential rank agreement under $\widetilde H_0$ is lower than the
sequential rank agreement under $H_0$ because the four methods used to
rank the data ($t$ test, logistic regression, elastic net, and MIC)
generally tend to identify the same predictors.
%even if there are only spurious associations.

\added{It is important to stress that neither $H_0$ nor
  $\widetilde H_0$ are related to questions regarding the association
  between the outcome and the predictors in the dataset. Both
  hypotheses are purely considering how the rankings agree in
  situations where the data contains no information for creating the
  rankings. It is also worth pointing out, that if the lists are short
  ($P$ low) and there are few lists ($L$ low) then the number of
  possible different permutations under the null is small and the $p$
  value obtained may be fluctuating if the number of permutations is
  small. We have found that a number of permutations over 500 works
  well even for smaller samples.}

@@

\begin{theorem}
  Assume that $\{R_l(X)\}_{l=1}^L$ are independent draws from a
  probability distribution $Q$ on the set of lists $\Pi$. Then
  $\left\|\widehat{\textrm{sra}}_L - \textrm{sra}\right\|_\infty =
  o_P(1)$.
\label{thm:consistency}
\begin{proof}
See App. A in the supplement.%ary% material.%\ref{sec:appA}.
\end{proof}
\end{theorem}

We now define the changepoint as the first crossing point of the
sequential rank agreement and a threshold function
$q\colon\,\{1,\ldots,P\} \mapsto \mathbb{R}_{\geq 0}$. The values of
$q$ could be a deterministic constant or, % in the experimental setting,
for example, the limits-of-agreement obtained in randomly permuted
lists corresponding to the null-hypothesis in equation
(\ref{eq:permutationHypothesis}). We define the superlevel set of the
sequential rank agreement with respect to $q$ as
\begin{align}
  \mathcal{L}(q) = \left\{d : \textrm{sra}(d) \geq q(d)\right\}.
\end{align}
A changepoint $d^\ast(q)$ in the list agreement is then defined by the position
\begin{equation}
  d^\ast(q) =
  \begin{cases}
    \inf(\mathcal{L}(q)) & |\mathcal{L}(q)| > 0\\
  P & |\mathcal{L}(q)| = 0
  \end{cases}
\end{equation}
corresponding to the first list depth where the sequential rank
agreement exceeds the threshold if such a position exists. Otherwise,
the full list is in agreement according to $q$ and the changepoint is
set to the full length of the lists. The empirical superlevel set is
similarly defined as
\begin{align}
 \widehat{\mathcal{L}}_L(\widehat{q}_L) = \left\{d : \widehat{\textrm{sra}}_L(d) \geq \widehat{q}_L(d)\right\}
\end{align}
where the threshold function may depend on the sample size.
% as well.
The estimated changepoint is% therefore
\begin{align}
  \widehat{d^\ast_L}(\widehat{q}_L) &= 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| > 0)\inf \widehat{\mathcal{L}}_L(\widehat{q}_L) + 1(|\widehat{\mathcal{L}}_L(\widehat{q}_L)| = 0)P.
\end{align}

The consistency of the estimated changepoint, $\widehat{d^\ast_L}(\widehat{q}_L)$, follows from Theorem \ref{thm:consistency} by the following corollary. 
\begin{corollary}\label{thecor}
Let $\widehat{q}_L$ be a positive threshold function such that $\left\|\widehat{q}_L - q\right\|_\infty = o_P(1)$ for some limiting
function $q$.  Then $\widehat{d^\ast_L}(\widehat{q}_L) \overset{P}{\longrightarrow} d^\ast(q)$ for $L \rightarrow \infty$.
\label{cor:consistency}
\begin{proof}
See App. B in supplement.%ary material.%\ref{sec:appB}.
\end{proof}
\end{corollary}

@@

The MALOVA study is a Danish study on ovarian cancer
\citep{Hogdall:2004:Cancer:15160342} where all Danish women diagnosed
with an ovarian tumor and referred for surgery from the participating
departments of gynecology were enrolled continuously from December
1994 to May 1999. For the purpose of illustration we use a random
sub-sample of $119$ patients with a total of $58$ patients with
malignant ovarian cancers as cases and $61$ patients with benign
ovarian tumors as controls. The DACOVA study is another
Danish study on ovarian cancer which included about
$66\%$ of the female population of Denmark
\citep{bertelsen1991protocol}. The study aimed to continuously enroll
all patients that were referred to surgery of an ovarian tumor
clinically suspected to be cancer during the period from 1984 to
1990. We use a random sub-sample from the DACOVA study of $54$
malignant ovarian cancers and $59$ benign ovarian tumors/gynecologic
disorders.

@@

In both scenarios we use the MALOVA data to train the statistical
models, and in both situations the agreements are assessed with
respect to perturbations of the training data in the following
manner. We draw 1000 random sub-samples (without replication)
consisting of 90\% of the MALOVA observations and train the four
models on each sub-sample.
%We use 1000 iterations for the sub-sampling
%procedure.

The implementation of Lasso and Ridge regression in the
\texttt{glmnet} package offers three different cross-validated
optimization criteria for the penalty parameter: total deviance,
classification accuracy and area under ROC. We apply all three
criteria to investigate their effect on the sra.
Note that these models produce incomplete lists depending on
the value of the penalty parameter.

@@

The top right panel of Figure \ref{fig:app1} shows the sequential rank
agreement of the estimated importance of the 5000 predictors.  For
clarity of presentation we zoom in on the agreement up to list depth
600. At deeper list depths all agreement curves are approximately constant.
%
As expected, most of the sequential rank agreement curves start low,
indicating good agreement, followed by an increase until they become
approximately constant. This has the interpretation that the
agreement across the sub-samples is higher in the top as
compared to the tail of the lists for all these classification
methods. The changepoints where the curves become approximately
constant are the list depths where the ranks of the remaining items
become close to uniformly random.

A not expected shape of the agreement curves is seen for the Ridge
models for all three tuning criteria. They all show higher
disagreement in the top of the lists followed by a decrease. The
reason behind this behavior is rather subtle. Looking at the
distribution of the absolute value of the regression coefficients we
see that a large proportion of them are numerically very close to zero
and have almost equal absolute value. This is a general feature of the
Ridge models in this dataset and seen for all the 1000 trained models.
This implies that when predictors are ranked according to the
magnitude of their coefficients, their actual order becomes more
uncertain and more close to a random permutation. This problem can be
alleviated by truncating all predictors with absolute coefficient
values below a given threshold thereby introducing an artificial
incompletion of the lists.  For the Ridge models tuned with the
deviance criterion, Figure \ref{fig:app1} (bottom left) shows the the
sequential rank agreement where for each of the 1000 trained models
the predictors were artificially censored when their absolute
coefficient value was lower than the $0.1\%$ quantile of the 5000
absolute coefficient values. The curve was calculated using Algorithm
1 from the appendix with $B=1000$ and $P=5000$. The corresponding
curve from Figure \ref{fig:app1} (top right) is shown for
comparison.  Even though the number of predictors with missing ranks is very
small compared to the total number of predictors, the effect on the
sequential rank agreement is substantial and with the artificial
censoring the shape of the curves is as expected, starting low and
then increasing.

Looking at the agreement curves for the Lasso models in Figure
\ref{fig:app1} (top right) we clearly see the effect of the sparsity
inducing penalization giving rise to incomplete lists. These curves
were similarly calculated using the algorithm and $1000$ random
permutations. Under the deviance optimization criterion the median
number of non-zero coefficients was 33 (range 16 to 50) and for the
class accuracy criterion 14 (range 4 to 56). These values correspond
to the list depths where the agreement curves become constant as a
result of the subsequent censoring.

\subsection{Agreement of individual risk predictions}
\label{sec:airp}
To assess the stability of the individual risk predictions we apply
the predictors from the DACOVA dataset to each of the models. The
predicted probabilities are then ranked in decreasing order such that
the patients with the highest risk of a malignant tumor appears in the
top of the list. Figure \ref{fig:app1} (top left) shows the sra separately
for each method, based on the 1000 risk predictions obtained from the
models trained in the same $1000$ random sub-samples of the MALOVA
data.
